---
title: "DSC1105 | FA 5"
author: "Baybayon, Darlyn Antoinette B."
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r library}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(MASS)
  library(caret)
  library(glmnet)
  library(cowplot)
  library(psych)
})
```


### Load and Explore the Data 

```{r}
sales_data <- read_csv("store_sales_data.csv", show_col_types = FALSE)
head(sales_data)
```


**Summary statistics and Variable types**

```{r}
summary(sales_data)
glimpse(sales_data)
```

The dataset, sales_data, contains 5,000 observations of simulated daily store sales. It has the following columns: \

- day_of_the week - categorical; 0,1,2,...,6 for each day of the week (Sunday-Saturday)
- promo - categorical; binary 1,0
- holiday - categorical;  binary 1,0
- store_size - categorical; levels: small, medium, large
- sales_count - numerical; no. of sales for the day \

Convert the columns to the appropriate data types.

```{r}
sales_data <- sales_data %>%
  mutate(across(c(day_of_week, promo, holiday, store_size), as.factor))

head(sales_data)
```

**Exploration of distribution of sales_count, store_size, and proportion of days with promo and holiday.**

```{r, fig.width=6, fig.height=3.5}
ggplot(sales_data, aes(x=sales_count)) +
  geom_histogram(binwidth = 10, fill="#F8766D") + 
  labs(title= "Distribution of Sales Count")

ggplot(sales_data, aes(x= store_size)) + 
  geom_bar(aes(fill=store_size)) + 
  labs(title="Frequency of Store Size")

sales_data %>%
  summarise(
    promo = mean(promo == 1),
    holiday = mean(holiday == 1)
  ) %>% pivot_longer(
    cols = c(promo, holiday), 
    names_to = "Type", values_to = "Proportion") %>%
  ggplot(aes(x = Type, y = Proportion, fill = Type)) +
  geom_bar(stat = "identity") +
  labs(title = "Proportion of Days with Promo and Holiday",
       x = "",
       y = "Proportion")
```


### Fit a Poisson Regression Model \

**Poisson Model**\
Take sales_count as the outcome and days_of_week, promo, holiday, and store_size as predictors.

```{r}
poisson_model <- glm(sales_count ~ day_of_week+promo+holiday+store_size, 
                     family=poisson, data = sales_data)

summary(poisson_model)
```

**Model Summary** \


*What happens to expected sales when there’s a promotion?* 

The sales count increases when there's a promotion. With a promotion, the expected sales count increases by a factor of $e^{0.410422}\approx1.507$.

*How does store size affect expected sales?*

Store size affects expected sales negatively. Smaller stores generate less sales. If a store is medium sized, its sales count would be about $e^{-0.697036}\approx0.498$ or 49.8% of the expected sales of large stores. Meanwhile, a small sized store would have $e^{-1.395325}\approx0.2478$ or 24.78% of the expected sales of large stores.


### Assess Model Fit \

Check for overdispersion

```{r}
deviance(poisson_model) / df.residual(poisson_model)
```
The ratio is 1.029, hence no significant overdispersion. This Poisson model is appropriate.

### Make Predictions \

Predict expected sales for a medium store on a Monday with a promotion and no holiday.

```{r}
exp(2.986855 + 0.060625 + -0.697036 + 0.410422 + 0) 

```

About 15.81 sales would be expected on a medium store on a Monday with a promo and no holiday. \

Predict sales for a large store on a Sunday with no promotion and a holiday.

```{r}
exp(2.986855 + 0 + 0 -0.330349)
```

About 14.25 sales would be expected on a Sunday with no promo and a holiday.

The second scenario expected less sales than the former since higher sales are usually seen on Monday-Saturday (positive coefficients) and holidays (negative coefficient) affect sales negatively.
\

### Reflection \

A poisson regression model was used to fit the data and predict daily sales count of stores with the following predictors: day of the week, store size, promotions, and holidays. To assess model fit, overdispersion was checked by calculating the ratio of residual deviance (5134.7) to the residual degrees of freedom (4989) resulting to a result of 1.03, indicating no significant overdispersion, and suggesting that the model provides an adequate fit to the data. Among the predictors, store size appears to have the strongest impact on expected sales. Small store size, in particular, has the largest coefficient of -1.395, which indicates substantial decrease in sales compared to larger stores. One of the primary limitations of the poisson distribution is its assumption that the mean is equal the variance. However, this is often not the case for real-world scenarios, where using poisson models can yield biased estimates. This model is also limited by the predictors available. Other potentially influential factors not included in the dataset, such as store location, is not explained by this model.




