---
title: "DSC1107 | FA 5"
author: "Baybayon, Darlyn Antoinette B."
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r library}
suppressPackageStartupMessages({
  library(tidyverse)
  library(dplyr)
  library(readr)
  library(ggplot2)
  library(stat471)
  library(caret)
  library(glmnet)
  library(cowplot)
  library(psych)
  library(pROC)
})
```


### Introduction

**The Dataset and Objectives**
```{r}
data <- read_csv("train.csv", show_col_types = FALSE)
head(data)
```

The dataset contains information about passengers of the Titanic and their survival status. It has 891 observations and 12 columns. The key features are as follows:

- Pclass - Passenger class (1st, 2nd, 3rd)
- Sex - Male or Female
- Age - Age in years
- SibSp - no. of Siblings/spouses aboard
- Parch - no. of parents/children aboard
- Fare - Passenger fare
- Embarked - Port of Embarkation
- Survived - Survival STatus


The objective of this project is to implement logistic regression to build a model that predicts a passenger's survival status based on the given information, analyze the model's performance, and interpret the results.


### Data Preprocessing

#### Exploratory Data Analysis

Dataset shape and data types
```{r}
glimpse(data)

```
**Statistical Summary**
```{r}
round(t(describe(data, omit=TRUE)),4)
```

Null
```{r}
sapply(data, function(x) sum(is.na(x)))
```

**Data Cleaning**

Drop irrelevant columns and convert categorical variables into numerical format by label encoding.

```{r}
data <- data %>%
  select(-PassengerId, -Name, -Cabin,-Ticket) %>%
  mutate(
    Sex = as.numeric(factor(Sex)),
    Embarked = as.numeric(factor(Embarked)),
    across(c(Survived, Pclass,Sex, Embarked), as.factor)
  )
  
head(data)
round(t(describe(data, omit=TRUE)),4)

```

**Data visualization**
```{r}
plot_grid(
  ggplot(data, aes(x=Fare)) + geom_histogram(binwidth = 40),
  ggplot(data, aes(x=Age)) + geom_histogram(binwidth = 10),
  ggplot(data, aes(x=SibSp)) + geom_histogram(binwidth = 1),
  ggplot(data, aes(x=Parch)) + geom_histogram(binwidth = 1),
  ncol= 2
)

plot_grid(
  ggplot(data, aes(x=Survived)) + geom_bar(),
  ggplot(data, aes(x=Sex)) + geom_bar(aes(fill= factor(Survived))) +
    labs(title="Survival by Sex"),
  ggplot(data, aes(x=Pclass)) + geom_bar(aes(fill= factor(Survived))) +
    labs(title="Survival by Passenger Class"),
  ggplot(data, aes(x=Embarked)) + geom_bar(aes(fill= factor(Survived))) +
    labs(title="Survival by Embarkation Point"),
  ncol=2
)

```

**Handle null values**

From the EDA, we found that Embarked column has 2 missing values and Age column has 177. Handle the missing values by imputation.

```{r}
data_clean <- data %>%
  mutate(
    Age = if_else(is.na(Age), median(Age, na.rm = TRUE), Age),
    Embarked = if_else(is.na(Embarked), factor(3), Embarked)
  )

```


```{r}
plot_grid(
  ggplot(data_clean, aes(x=Fare)) + geom_histogram(binwidth = 40),
  ggplot(data_clean, aes(x=Age)) + geom_histogram(binwidth = 10),
  ggplot(data_clean, aes(x=SibSp)) + geom_histogram(binwidth = 1),
  ggplot(data_clean, aes(x=Parch)) + geom_histogram(binwidth = 1),
  ncol= 2
)

plot_grid(
  ggplot(data_clean, aes(x=Survived)) + geom_bar(),
  ggplot(data_clean, aes(x=Sex)) + geom_bar(aes(fill= factor(Survived))) +
    labs(title="Survival by Sex"),
  ggplot(data_clean, aes(x=Pclass)) + geom_bar(aes(fill= factor(Survived))) +
    labs(title="Survival by Passenger Class"),
  ggplot(data_clean, aes(x=Embarked)) + geom_bar(aes(fill= factor(Survived))) +
    labs(title="Survival by Embarkation Point"),
  ncol=2
)

```


**Splitting the dataset**

Split dataset into training and testing set (80-20 split).
Oversample the minority class (survivors) to address class imbalance.

```{r}
set.seed(123)
table(data_clean$Survived)

data_clean <- upSample(x = data_clean[, -1], y = data_clean$Survived)
names(data_clean)[ncol(data_clean)] <- "Survived"
table(data_clean$Survived)

```


```{r}

n <- nrow(data_clean)
train_samples <- sample(1:n, round(0.8*n))

data_train <- data_clean[train_samples, ]
data_test <- data_clean[-train_samples, ]

```



### Model Implementation

**Training the model on the training set**
```{r}
model_logistic <- glm(Survived ~ Pclass + Sex + Age + SibSp + Parch + Fare + Embarked, 
                      data=data_train, family="binomial")
summary(model_logistic)

```

**Predict outcomes on the test set**
```{r}

predictions <- predict(model_logistic, newdata = data_test, type="response")
prediction_class <- factor(if_else( predictions > 0.5, 1, 0))

```


### Model Evaluation
**Performance metrics**

```{r}
conf_matrix <- confusionMatrix(data_test$Survived, prediction_class)$table

cat("Confusion Matrix:\n")
print(conf_matrix)
TN <- conf_matrix[1,1]
FP <- conf_matrix[1,2]  
FN <- conf_matrix[2,1]  
TP <- conf_matrix[2,2]  

accuracy <- (TP + TN) / sum(conf_matrix)
precision <- TP / (TP + FP)
recall <- TP / (TP + FN)
f1_score <- 2 * (precision * recall) / (precision + recall)


data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F-1 Score"),
  Value = round(c(accuracy, precision, recall, f1_score),4)
)

```
```{r}
confusionMatrix(data_test$Survived, prediction_class)
```

ROC Curve and AUC score
```{r}
roc_curve <- roc(data_test$Survived, predictions)
plot(roc_curve, col = "blue", main = "ROC Curve", print.auc = TRUE)

abline(a = 0, b = 1, lty = 2, col = "red")

```


### Results Interpretation and Discussion

**Summary**

```{r}
summary(model_logistic)
```


A logistic regression model was trained to predict the likelihood of survival of passengers of the Titanic based on Passenger Class (Pclass), Sex, Age, No. of Siblings/Spouses aboard (SibSp), No. of Parents/Children aboard (Parch), Fare, and Port of Embarkation (Embarked). Among these features, Pclass, Sex, Age, and SibSp were found to be statistically significant (p < 0.05).

The logistic regression model exhibited good overall performance, with an accuracy of 78.18%, precision of 78.70%, recall of 77.27%, and an F1 score of 77.98%. The AUC 
(Area under Curve) of the ROC (Receiver Operating Characteristic) was 0.839, which indicates the model's strong ability to distinguishing between survivors and non-survivors.


**Significant Features and interpretation**

Pclass2 (B= -1.159, p = 0.000162) had lower odds of survival compared to Pclass1; Pclass3 (B = -2.197, p = 8.97e-13) had lower odds of survival compared to class1m with a stronger impact than Pclass2. These resuls indicate that passengers in lower classes were more unlikely to survive.

Sex2 (B = -2.648, p < 2e-16) had a strong impact on survival status, with males being significantly less likely to survive than females.

Age (B = -0.046, p = 1.10e-08) was also negatively correlated with survival, decreasing the likelihood of survival as age increases.

SibSp (B = -0.47, p = 0.000118) also negatively impacted survival, which indicate that individuals in larger families struggled to survive.


**Recommendations**

To further improve the performance of this model, some strategies can be implemented. Feature engineering can help create new predictors. For example, SibSp and Parch may be combined to a new variable, FamilySize.



